GEE models

### 2017-07-20 doc/submission/reviewer-response.md

# Responses to reviewers

## Referee: 1

- *"First, I would like to congratulate you for the exhaustive and complete
research work that has been done for this paper. I think the statistical study
is very successful and useful for researchers who have accumulated a large
database and need a method with the ability to handle a large number of
variables and their interactions. I really liked read in the discussion the
negative points of the study, with which I fully agree, and not an attempt to
hide the limitations of any scientific work. In retrospect, perhaps it will miss
data that clarify the pre-diabetic state, or not, of the patients (HbA1c or
C-peptide). I think it would be good for the study know if some patients were
taking any medication that could affect the results. Although it is a
complicated scientific work to understand for researchers who do not usually use
this kind of statistical studies, I believe that is described in a very
appropriate and clear manner."*

    We thank the reviewer for these comments. In the PROMISE cohort, we
    collected medication data from all participants. While most participants
    took some form of medication over the 6 years, during the times of data
    collection, many (n=60-103 between the three visits) were on taking lipid
    and cholesterol controlling medications. However, for those who were taking
    medications that could influence the results, additional analyses revealed
    that medication use did not substantially influence the predictor or outcome
    measures, nor did medication use substantially contribute to GEE model fit
    compared to models without it (see QIC table). We have included this
    information within the manuscript.

@@

2. *"Methods in the abstract: a) Please give some information on the cohort
characteristics: age, sex, high risk of diabetes but no prevalent disease at
baseline; b) incident diabetes cases (how were these handled? what about treated
patients? - these should go in the main methods section); c) adjustment for time
– I would prefer to state that time since baseline was used as the underlying
time variable instead of adjusting for time"*

    We have made as many of the requested changes to the methods as the reviewer
    suggested without going over the word limit. Incident diabetes cases are
    presented for descriptive purposes though were not included in any GEE
    analysis. We thank the reviewer for the suggestion regarding time adjustment
    and agree that using time since baseline is a better variable to adjust for
    in the GEE models; the time variable has been replaced in GEE models, though
    the change does not substantially impact the final results.

3. *"Results in the abstract: a major problem in my view with the
interpretation of the results is that all characteristics are taken as
confounders although a considerable proportion of time-varying covariates are
taken after the NEFA measurements. Furthermore, it is not clear whether these
measures are confounders, mediators or NEFAs are mediators of these variables. I
think that unadjusted results are as interesting and require discussion as the
fully adjusted and selected models. Regarding the abstract, I would show both
the unadjusted and the fully adjusted results here."*

    The specifics of the covariates are expanded upon in the methods section. We
    have included the unadjusted results in the main manuscript and have 
    described it in the results and discussion sections, in addition to briefly
    describing them in the abstract (keeping within the word limit). An
    explanation of what each covariate is in regard to NEFA and the outcomes
    (e.g. mediator or confounder) has been added in the ESM Methods.

@@

    - *"I am not an expert of PLS-DA but it is not clear to me, why does this
    analysis reflect more the role of relative NEFA (mole%) analysis than the
    concentration analysis? As I can see, it depends on what variables you feed
    in the model."*

        We are not entirely clear on the reviewer's comment. PLS extracts from
        high dimensional data (multiple predictor variables, i.e. the fatty
        acids) underlying correlations, constrained by a outcome variable (i.e.
        beta-cell function). In this context, because mol% data is by definition
        restricted to sums of 100% (all fatty acid values add up to 100%), the
        underlying correlations between fatty acids is dependent on the relative
        contributions those fatty acids have to the total and as such can
        reflect correlations of specific fatty acids or groups of fatty acids
        that increase or decrease in proportion together. The concentration data
        is not contrained by this inherent correlation and thus underlying groups 
        or clusters may not be identified using PLS.

7. *"RESULTS:"*
    - *"Please give ns in Table 1."* ... *"It would be also nice to see, how
    many participants had 1, 2 or 3 time points in the analysis."*

        These have been added as requested. {{Tony, not sure about how I feel
        adding signif to Table 1. My analysis isn't on changes in these
        measures, so there is not need to test if there is a difference. It's
        purely descriptive and I'd prefer not to generate p-values for the sake
        of generating p-values. Thoughts? I could always just do a baseline table...}}

@@

    - *"I suggest to include the unadjusted data to the main paper (ESM Figure 3)
    and describe these findings as we do not know the exact model where
    over-adjustment start to happen."*

        The unadjusted results and figure has been added to the main paper and
        described in the results section. We've also added a sensitivity
        analysis confirming which covariate(s) attenuates the associations.

    - *"In a trajectory analysis, the interaction between time and the main
    predictor (NEFA in this case) is of major interest. The authors check for
    this interaction only as the last step of their model building (see ESM). I
    would suggest to perform this as the first step of the analysis (M0), and
    drop this term, if it is non-significant or does not improve model fit.
    Furthermore, given that this interaction was dropped from the model, it is
    hard to except the major conclusion that NEFA predicts progression of
    beat-cell dysfunction. Actually what is found here is that NEFA are related
    to beta-cell function in repeated cross-sectional analyses and NEFA related
    differences in beta-cell function remain constant over time."*

        The reviewer makes an excellent point regarding the importance of
        conducting an interaction check in the early stages of
        model building. We have incorporated this new approach in the analysis and
        report on the results in the paper. In simple terms, the reviewer is
        correct that without an interaction term the model is essentially
        "repeated cross-sectional analyses", however, GEE handles the analysis
        in a slightly more nuanced approach. GEE calculates the estimates of the
        association of NEFA on the outcomes at any given time point, taking into
        consideration the inherent multivariate structure of the repeated
        measurements of the outcome variables. As such, inherent similarities in
        the outcome measurements over time within a subject are included in the
        computation of the GEE estimates. This nuance provides substantially
        more power to the model and the proceeding results. See ref 27 in the
        manuscript for more detail on this statistical technique. Within the
        manuscript we have expanded on this nuance to clarify the
        interpretation.

@@

    - *"Latent classes. Although this is a very attractive method, based on ESM
    figure 4, there are 3 parallel declining group trajectories that is mostly
    affected by the baseline values. Groups based on baseline (last follow-up)
    tertiles would be easier to interpret."* ... *"The PLS-DA analysis needs
    further description. I would expect variables with the highest loadings to
    be the variables in the unadjusted GEE analysis. I am probably wrong given
    the results."* ... *"Do we really expect good discrimination? The
    associations are week to nul between NEFAs and the outcomes. I do not really
    see the point the extensive supplementary analysis on PLS-DA. Furthermore
    dysglycemia is an outcome that is not directly related to beta-cell function
    only. Dysglycemia requires both insulin resistance and beta-cell
    dysfunction."*

        Given this reviewer's previous comments and suggestions, we have decided
        to: focus on the GEE analysis in the main paper; remove the LCMM
        analysis; perform the PLS analysis on the baseline (cross-sectional)
        data using the outcomes as continuous variables; and, briefly describe
        the findings in the Results section. To keep the ESM material
        streamlined, we have contained the PLS results and other additional
        analyses conducted within the code (accessible via the DOI) for the
        interested reader. We believe this simplifies the results and primary
        findings of this paper without sacrificing a loss in the potential use
        these additional analyses may provide for the interested researcher.

    - *"The authors mention in the Discussion the external validity is mostly
    limited to white females. I suggest to run a sensitivity analysis on this
    group."*

        We agree with the reviewer's comment and ran stratified models by either
        sex or ethnicity. In both cases, all significant associations from the
        non-stratified analyses were attenuated, likely due to the reduction in
        sample size in the stratified models. Stratifying analyses has the
        limitation of substantially reducing sample size and is the reason why
        interaction testing is prefered. We ran interaction tests by sex and
        ethnicity and found no significant interactive associations. While
        these results internally suggest no differences between sex or ethnicity,
        there may still be features unique to sex and ethnicity in this cohort
        that limit generalizability.

### 2017-07-20 doc/esm-methods.Rmd
added info on what each covariate is in esm methods
cat for formulas for proper printing

```{r}
PROMISE.methods::insert_formula("ISI")
```

```{r}
PROMISE.methods::insert_formula("HOMA")
```

```{r}
PROMISE.methods::insert_formula("IGIIR")
```

```{r}
PROMISE.methods::insert_formula("ISSI2")
```

@@

For the covariates, they were chosen based on previous literature, from 
directed acyclic graph (DAG) recommendations, and from QIC. The DAG
recommendations were obtained from using the DAGitty software,
http://dagitty.net/. DAG is a technique that allows for identifying potential
confounding causal pathways and to limit potential bias introduced from
adjusting for colliding variables. A collider variable is a variable that is
determined from two or more other variables, such that in causal graph modeling,
the arrows from the other variables appear to 'collide' into the variable. QIC 
is a technique that balances the goodness of fit of the model with its relative
complexity (i.e. number of covariates) against other models being compared. This
identifies a specific model from a number of models that is the 'best' fit and
least complex of all the models. QIC and other information criterion methods are
used in conjunction with other techniques such as literature and DAG based. The 
final GEE model selected as best fitting was negligibly different when using 
insulin sensitivity or beta-cell function as an outcome and as such the same 
covariates were chosen for models with each of the outcomes.

>>>>

```{r}
cat(PROMISE.methods::insert_formula("ISI"))
```

```{r}
cat(PROMISE.methods::insert_formula("HOMA"))
```

```{r}
cat(PROMISE.methods::insert_formula("IGIIR"))
```

```{r}
cat(PROMISE.methods::insert_formula("ISSI2"))
```

@@

For the covariates, they were chosen based on previous literature, from 
directed acyclic graph (DAG) recommendations, and from QIC. The DAG
recommendations were obtained from using the DAGitty software,
http://dagitty.net/. DAG is a technique that allows for identifying potential
confounding causal pathways and to limit potential bias introduced from
adjusting for colliding variables. A collider variable is a variable that is
determined from two or more other variables, such that in causal graph modeling,
the arrows from the other variables appear to 'collide' into the variable. QIC 
is a technique that balances the goodness of fit of the model with its relative
complexity (i.e. number of covariates) against other models being compared. This
identifies a specific model from a number of models that is the 'best' fit and
least complex of all the models. QIC and other information criterion methods are
used in conjunction with other techniques such as literature and DAG based. The 
final GEE model selected as best fitting was negligibly different when using 
insulin sensitivity or beta-cell function as an outcome and as such the same 
covariates were chosen for models with each of the outcomes. The covariates
medication use, WC, ALT, and MET would be considered classical confounders while
TAG would be considered a mediating variable.
